---
title: "Lab 02 - Plastic waste"
author: "Saima Arina"
date: "01/22/2026"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

### Exercise 1

In terms of their plastic waste per capita, it seems that North America has the highest, with the Trinidad and Tobago outlier of over 3.5 kg/person and also a skew towards higher numbers. Oceania and South America share similar distributions. Africa seems to have the lowest, with most countries below 0.2 kg/person. Asia seems to have a skew towards the right. 

```{r plastic-waste-continent}
ggplot(data = plastic_waste, mapping = aes (x = plastic_waste_per_cap)) + geom_histogram (binwidth = 0.1) +
facet_wrap(~ continent)
```

### Exercise 2

```{r plastic-waste-density}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent,
    fill = continent,
  )
) +
  geom_density(alpha = 0.175)
```

The color and fill of the curves were defined by mapping aesthetics because if the info is to be in the legend, it must be a part of this layer. Also, it allows the colors to scale for each group. The alpha level is defined as a characteristic of the plotting geom because it is a element universally applied to all of the groups rather than differently for them individually.

### Exercise 3

Remove this text, and add your answer for Exercise 3 here.

```{r plastic-waste-violin}
# insert code here
```

### Exercise 4

Remove this text, and add your answer for Exercise 4 here.

```{r plastic-waste-mismanaged}
# insert code here
```

```{r plastic-waste-mismanaged-continent}
# insert code here
```

```{r plastic-waste-population-total}
# insert code here
```

```{r plastic-waste-population-coastal}
# insert code here
```

### Exercise 5

Remove this text, and add your answer for Exercise 5 here.

```{r recreate-viz}
# insert code here
```


