---
title: "Lab 02 - Plastic waste"
author: "Saima Arina"
date: "01/22/2026"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

### Exercise 1

In terms of their plastic waste per capita, it seems that North America has the highest, with the Trinidad and Tobago outlier of over 3.5 kg/person and also a skew towards higher numbers. Oceania and South America share similar distributions. Africa seems to have the lowest, with most countries below 0.2 kg/person. Asia seems to have a skew towards the right. 

```{r plastic-waste-continent}
ggplot(data = plastic_waste, mapping = aes (x = plastic_waste_per_cap)) + geom_histogram (binwidth = 0.1) +
facet_wrap(~ continent)
```

### Exercise 2

```{r plastic-waste-density}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent,
    fill = continent,
  )
) +
  geom_density(alpha = 0.175)
```

The color and fill of the curves were defined by mapping aesthetics because if the info is to be in the legend, it must be a part of this layer. Also, it allows the colors to scale for each group. The alpha level is defined as a characteristic of the plotting geom because it is a element universally applied to all of the groups rather than differently for them individually.

### Exercise 3

Violin plots are capable of revealing the shape of the distribution, including peaks and skewness, while box plots provide more information on quartiles and extreme outliers. 

```{r plastic-waste-violin}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = continent,
    y = plastic_waste_per_cap
  )
) +
  geom_boxplot()
```

```{r}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = continent,
    y = plastic_waste_per_cap
    
  )
) +
  geom_violin()
```


### Exercise 4

There is a pretty visible positive linear relationship where higher total plastic waste per capita is associated with higher amounts of mismanaged plastic waste per capita, and most points are clustered along a line that is upwards and there are some high outliers. In terms of continent, all of the colors seem to flow the same positive trend, with North America showing prominently high-value outliers. I would say for the last two plots, that plastic waste per capita and total population seem to be more strongly linearly associated because more of the data points are clustered together in the same area while the other plot measuring coastal population has a greater numbers of outliers and slightly more spread. 

```{r plastic-waste-mismanaged}
ggplot(data = plastic_waste,
       mapping = aes(x = plastic_waste_per_cap,  y = mismanaged_plastic_waste_per_cap
                
                     )) +
  geom_point() 
                   
```

```{r plastic-waste-mismanaged-continent}
ggplot(data = plastic_waste,
       mapping = aes(x = plastic_waste_per_cap,  y = mismanaged_plastic_waste_per_cap,
                     color = continent
                     )) +
  geom_point() 
                   
```

```{r plastic-waste-population-total}
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, y = total_pop)) + geom_point ()
```

```{r plastic-waste-population-coastal}
ggplot(data = plastic_waste, 
       mapping = aes (x = plastic_waste_per_cap, y = coastal_pop)) + geom_point ()
```


### Exercise 5


```{r recreate-viz}
ex5data = mutate(filter(plastic_waste,plastic_waste_per_cap <= 3),
         coastal_prop = coastal_pop / total_pop) 

  ggplot(data = ex5data,
    aes(
      x = coastal_prop,
      y = plastic_waste_per_cap,
      color = continent
    )
  ) +
  geom_point(alpha = 0.4) +
  geom_smooth(
    method = "loess",
    se = TRUE,
    color = "black"
  ) +
  scale_color_viridis_d() +
  labs(
    title = "Plastic waste vs. coastal population proportion",
    subtitle = "by continent",
    x = "Coastal population proportion (Coastal / total population)",
    y = "Plastic waste per capita",
    color = "Continent"
  ) +
    scale_color_viridis_d()
```


I had to look some of this code up, because I got stuck trying to figure out how to code for things like the regression line and how to do the shading around it, but then I plugged it into AI and asked it to explain the code one by one and it makes more sense. 

Overall, I see that countries with a moderate share of their population living in coastal areas tend to have higher plastic waste per capita than those at either very low or very high coastal proportions, showing a nonlinear association instead of a more simple straight-line trend. At low coastal population proportions, plastic waste per capita is generally sized, then increases as the coastal share rises, before tapering off again among countries where almost everyone lives near the coast. 